---
title: "R Notebook"
output: html_notebook
---

This needs the following packages installed:
```{r}
install.packages(c(
  "DBI",
  "RPostgres",
  "dplyr",
  "dbplyr",
  "ggplot2",
  "knitr",
  "purrr",
  "rmarkdown",
  "stringr",
  "tibble",
  "glue"
))
```

```{r, results='hide', warning=FALSE}
library(DBI)
library(dplyr)
library(ggplot2)
library(tibble)
library(glue)
library(stringr)
library(purrr)
```

launch the postgres database in a docker container:
```{bash}
docker-compose up -d
```

# TODO: look into https://serverfault.com/a/935674 for code that checks if container is up

make a connection to the database
```{r}
con <- dbConnect(
    RPostgres::Postgres(),
    dbname = "spirographs",
    host = "localhost",
    port = 5432,
    password = "postgres",
    user = "postgres"
)
```

create the database
```{sql create-db-with-xy-cols-as-int-arrays, connection=con}
CREATE TABLE cached_runs_2 ( 
  n1 integer,
  n2 integer,
  n3 integer,
  x integer[], 
  y integer[]
)
```

```{r spiro-without-db}
spiro_without_db <- function(n1,n2,n3) {
  # calculate the new spirograph points
  t <- seq(0,1,length.out=1000)
  z <- exp(1i*2*pi*n1*t) + exp(1i*2*pi*n2*t) + exp(1i*2*pi*n3*t)
  result <- tibble(x=Re(z),y=Im(z))
  return (result)
}
```

add a run to the db.
1. use the array syntax described in [documentation section 8.15.2](https://www.postgresql.org/docs/current/arrays.html#ARRAYS-INPUT).
2. use `glue_sql()` to properly quote the sql statement. note that we change `glue`'s default open and close markers from `{` and `}` to `<<` and `>>` so we can use the curly brackets in the sql statement to signify an array input.
3. use the `*` to [join the list elements with commas](https://glue.tidyverse.org/reference/glue_sql.html)
```{r save-run-to-db}
n1 <- 0
n2 <- 0
n3 <- 0
r <- spiro_without_db(n1, n2, n3)

query <- glue_sql("
  INSERT INTO cached_runs_2
  VALUES (
    <<n1>>,
    <<n2>>,
    <<n3>>,
    '{<<r$x*>>}',
    '{<<r$y*>>}'
  );
  ", .con = con, .open="<<", .close=">>")

query

q <- dbExecute(con, query)
```

get the list data from db as a list
```{r}
query <- glue_sql("SELECT * FROM cached_runs_2")
results <- dbGetQuery(con, query)
results
```

convert the x and y columns to an double arrays. similar solutions in https://stackoverflow.com/q/40637545

```{r convert-result-x-y-character-cols-to-double-list}
as.numeric_list <- function(x) {
  result <- x %>%
    str_sub(2,-2) %>%    # remove the { and } characters
    str_split(",") %>%   # split the string based on comma
    unlist %>%           # remove one level of lists that gets added by str_split()
    map_dbl(as.double)   # convert each element to a double,
                         # use map_dbl so it will return an atomic vector of doubles
  
  return(result)
}

x <- as.numeric_list(results$x)
y <- as.numeric_list(results$y)

tribble(
  ~x, ~y,
  x, y
)
```

PostgreSQL supports other datatypes. You can see them listed in the [documentation](https://www.postgresql.org/docs/9.3/datatype-geometric.html). We could try using one of the geometric datatypes to store the x and y data points.

Let's try storing the data as an open path.

First create the database table:
```{sql create-db-table-path, connection=con}
CREATE TABLE cached_runs_3 (
  n1 integer,
  n2 integer,
  n3 integer,
  points path
)
```

```{r save-run-to-db-as-path}
n1 <- 0
n2 <- 0
n3 <- 0
r <- spiro_without_db(n1, n2, n3)

# format the points as a string like:
# "(x1,y1), ..., (xn,yn)"
points_str <- map2_chr(
    r$x, 
    r$y,
    function(x,y) {glue("({x},{y})")}
  ) %>%
  str_c(collapse = ",")

# surround the string with square brackets
# do this before the glue_sql so we can get the
# quotes outside the square brackets in the sql statement
points_str <- glue("[{points_str}]")

# craft the sql statement
query <- glue_sql("
  INSERT INTO cached_runs_3
  VALUES ( {n1}, {n2}, {n3}, {points_str} );
  ", .con = con)

query

dbExecute(con, query)
```

get the list data from db as a list
```{r}
query <- glue_sql("SELECT * FROM cached_runs_3")
results <- dbGetQuery(con, query)
results
```

To use this value in a plot, we have to parse the x and y values from the path and put them into separate columns of a data frame.

One advantage of geometric fields is that we can use postgresql's built in geometric functions. For more information on on the geometric functions see https://www.postgresql.org/docs/current/functions-geometry.html


disconnect
```{r}
dbDisconnect(con)
```

shut down the docker container:
```{bash}
docker-compose down
docker volume rm spiro_with_postrgresql_pg_data
```