---
title: "Leaflet & Plotly Mapping Examples"
output:
  html_document: default
  html_notebook: default
---

```{r echo=FALSE, results='hide'}
# setup file paths
base.dir = '.'
data.dir = file.path(base.dir,'data')
```


Data from a file CSV files can be read into a dataframe and fed into the Leaflet interface

```{r}
# indiana precipitation dataframe
stations.df <- read.csv(
                 file.path(
                   data.dir,'indiana_precipitation/station.csv'))
print(head(stations.df))
```

The data has fields named `latitude` and `longitude` which makes it easy for the leaflet library to locate the columns for latitude and longitude. Let's plot the locations and setup a label for each point.

First, create a `leaflet()` map object and set it up to use the default tile source from OpenStreetMap.

```{r}
# indiana precipitation map
ipm <- leaflet() %>%
         addTiles()
```

Next, add markers at each station's location using the `addMarkers()` function.

```{r}
ipm %>% addCircleMarkers(
          data = stations.df,
          lng = ~longitude,
          lat = ~latitude,
          radius = 4,
          layerId = ~Station.ID,
          fillOpacity = 0.8,
          popup = paste(
            "Station Name: ", stations.df$Station.Name, "<br>",
            "Station ID: ", stations.df$Station.ID, "<br>",
            "Latitude: ", stations.df$latitude, "<br>",
            "Longitude: ", stations.df$longitude, "<br>"
          )
        )
```

Read in the precipitation data for each station

```{r}
precip.df = data.frame()
precip.dir = file.path(data.dir,"indiana_precipitation")
precip.files = list.files(path=precip.dir,pattern="hr.*.csv")
for (i in 1:length(precip.files)) {
    matches = gregexpr("[0-9]+",precip.files[i])
    station.id = regmatches(precip.files[i],matches)[[1]]
    p = file.path(precip.dir, precip.files[i])
    df = read.csv(p)
    df$Station.ID = station.id
    precip.df = rbind(precip.df,df)
}
print(head(precip.df))
```


Visualize the precipitation data for station id 120132

```{r}
station.df = precip.df[precip.df$Station.ID == "120132",]

library(ggplot2)

p = ggplot(station.df,aes(Time,Precipitation)) +
    geom_line()
```

Visualize the plot using plotly, a service that provides javascripty d3.js plots

```{r}
library(plotly)

ggplotly(p)
```
