---
title: "Shiny Mapping Examples"
output:
  html_notebook: default
  runtime: shiny
---

Let's briefly recap what our data looks like. 

```{r, echo=FALSE, result='hide'}

base.dir = '.'
data.dir = file.path(base.dir,'data')
precip.dir = file.path(data.dir,'indiana_precipitation')

# read in location and description of the precip monitor stations
stations.df <- read.csv( file.path(precip.dir,'station.csv') )

# read in the precipitation data
precip.df = data.frame()
precip.files = list.files(path=precip.dir,pattern="hr.*.csv")

for (i in 1:length(precip.files)) {
  # get the station id from the file path
  matches = gregexpr("[0-9]+",precip.files[i])
  station.id = regmatches(precip.files[i],matches)[[1]]

  # save the file to our dataframe,
  # associating the station id with each row
  p = file.path(precip.dir,precip.files[i])
  df = read.csv(p)
  df$Station.ID = station.id
  precip.df = rbind(precip.df,df)
}

```

Stations is a dataframe with the longitude and latitude location of the observation sites. It also includes a station id and a station name:
```{r}
print(head(stations.df))
```

Precipitation is a dataframe holding all of the precipitation counts for each observation station:
```{r}
print(head(precip.df))
```

We can combine our leaflet map with a plotly graph. When a user clicks a marker on the map, the plot will be automatically updated with precipitation values.

```{r, echo=FALSE, result='hide'}
library(dplyr)
library(ggplot2)
library(lattice)
library(leaflet)
library(shiny)
library(plotly)
```

```{r, echo = FALSE}

# uses the following dataframes from above:
# stations.df
# precip.df

shinyApp(

  ui = bootstrapPage(
    tags$style(type = "text/css", "html, body {width:100%;height:100%}"),
    leafletOutput("map", width = "100%", height = "100%"),
    absolutePanel (bottom = 15, right = 2,
                  plotOutput("plot.lattice", height=250, width=300),
                  plotlyOutput("plot.plotly", height=250, width=300)
    )
  ),

  server = function(input, output) {
    data <- reactiveValues(
      clickedMarker=NULL,
      station.df=data.frame(Time=c(0),Precipitation=c(0))
    )

    # save information about which marker was clicked
    observeEvent(input$map_marker_click, {
      p <- input$map_marker_click
      data$clickedMarker <- p

      # lookup the station id of the selected marker
      # could have also filtered by p$id if we set the
      # layerId parameter in the addCircleMarkers() function.
      station.id <- (stations.df %>%
                       filter(latitude == p$lat,
                              longitude == p$lng) %>%
                       select(Station.ID)
                    )[[1]]

      # create a Time/Precip dataframe for the selected marker
      # this will be used later to generate our plot
      data$station.df <- precip.df %>%
                           filter(Station.ID == station.id)
    })

    # clicks on non-marker items do nothing
    observeEvent(input$map_click, {
      data$clickedMarker <- NULL
    })

    # print out info for the clicked marker
    output$clickInfo <- renderPrint({data$clickedMarker})

    # generate our Leaflet based map output 
    output$map <- renderLeaflet({
      leaflet() %>%
        addTiles() %>%
        addCircleMarkers(
          data = stations.df,
          lng = ~longitude,
          lat = ~latitude,
          radius = 4,
          fillOpacity = 0.8,
          layerId = ~Station.ID,
          popup = paste(
            "Station Name: ", stations.df$Station.Name, "<br>",
            "Station ID: ", stations.df$Station.ID, "<br>",
            "Latitude: ", stations.df$latitude, "<br>",
            "Longitude: ", stations.df$longitude, "<br>"
          )
        )
    })

    # generate our lattice based xy curve
    # based on the data in our reactive dataframe data$station.df
    output$plot.lattice <- renderPlot({
      if (is.null(data$station.df)) {
        return(NULL)
      }
      print(xyplot(Precipitation ~ Time, data = data$station.df))
    })

    # generate our Plotly based xy curve
    # based on the data in our reactive dataframe data$station.df
    output$plot.plotly <- renderPlotly({
      if (is.null(data$station.df)) {
        return(NULL)
      }

      # using the plot_ly() function (it's slow!)
      #plot_ly(data$station.df, x = ~Time, y = ~Precipitation)

      # using the ggplotly() function (it's faster)
      p = ggplot(data$station.df,aes(Time,Precipitation)) +
          geom_line()
      ggplotly(p)
    })

  },

  options = list(height = 500)
)
```


