---
title: "mtcars in a PostgresSQL Database"
output: html_notebook
---

This needs the following packages installed:
```{r}
install.packages(c(
  "DBI",
  "RPostgres",
  "dplyr",
  "dbplyr",
  "ggplot2",
  "knitr",
  "rmarkdown"
))
```

```{r}
library(DBI)
library(dplyr)
library(ggplot2)
```

launch the postgres database in a docker container:
```{bash}
docker-compose up -d
```

# TODO: look into https://serverfault.com/a/935674 for code that checks if container is up

make a connection to the database
```{r}
con <- dbConnect(
    RPostgres::Postgres(),
    dbname = "examples",
    host = "localhost",
    port = 5432,
    password = "postgres",
    user = "postgres"
)
```

copy the mtcars data to the database
details from: https://www.datacareer.de/blog/connect-to-postgresql-with-r-a-step-by-step-example/
```{r}
data('mtcars')
mtcars.df <- data.frame(carname = rownames(mtcars), mtcars, row.names = NULL)
mtcars.df$carname <- as.character(mtcars.df$carname)
copy_to(con, mtcars.df, name="mtcars", temporary=FALSE, overwrite=TRUE)
```

use `dbplyr` to lazily retrieve data:
notes from: https://dbplyr.tidyverse.org/
```{r}
mtcars2 <-tbl(con,"mtcars")
mtcars2
```

make a plot using data from the database:
notes from: http://michaeltoth.me/a-detailed-guide-to-the-ggplot-scatter-plot-in-r.html
```{r}
ggplot(mtcars2) +
  geom_point(aes(x=wt, y=mpg))
```

make a query to the database:
notes from: https://dbplyr.tidyverse.org/

```{r}
mtcars3 <- mtcars2 %>%
  filter(wt >= 3, wt <= 4)
```

show the sql query that would be executed:
```{r}
mtcars3 %>% show_query()
```

execute the query, i dont think this is necessary to make the following plot
```{r}
mtcars3 %>% collect()
```

plot the subset of mtcars
```{r}
ggplot(mtcars3) +
  geom_point(aes(x=wt, y=mpg))
```

disconnect
```{r}
dbDisconnect(con)
```

shut down the docker container:
```{bash}
docker-compose down
docker volume rm bank_example_pg_data
```