---
title: "mtcars in a SQLite Database"
output: html_notebook
---

This needs the following packages installed:
```{r install-packages}
install.packages(c(
  "DBI",
  "RSQLite",
  "dplyr",
  "dbplyr",
  "ggplot2",
  "knitr",
  "rmarkdown"
))
```

```{r load-packages}
library(DBI)
library(dplyr)
library(ggplot2)
```

For this example, we will use an "in-memory" database:
```{r create-in-memory-database}
con <- dbConnect(RSQLite::SQLite(), ":memory:")
```

One way to copy the mtcars data to the database is by using the `copy_to()` function.
With this code, we add an extra column with the names of the cars.
```{r write-dataframe-to-database}
data('mtcars')
mtcars.df <- data.frame(carname = rownames(mtcars), mtcars, row.names = NULL)
mtcars.df$carname <- as.character(mtcars.df$carname)
copy_to(con, mtcars.df, name="mtcars", temporary=FALSE, overwrite=TRUE)
```

Or you can use the more traditional r-dbi function `dbWriteTable()`.
This method doesn't include the column with the car names.
```{r another-way-write-dataframe-to-database}
dbWriteTable(con, "mtcars-without-names", mtcars)
dbListTables(con)
```

use `dbplyr` to lazily retrieve data:
```{r retrieve-data-with-dbplyr}
mtcars2 <- tbl(con, "mtcars")
mtcars2
```

make a plot using data from the database:
```{r plot-data-from-database}
ggplot(mtcars2) +
  geom_point(aes(x=wt, y=mpg))
```

Make a query to the database using dbplyr:
```{r query-database-with-dbplyr}
mtcars3 <- mtcars2 %>%
  filter(wt >= 3, wt <= 4)
```

show the sql query that would be executed:
```{r show-dbplyr-query-as-sql}
mtcars3 %>% show_query()
```

execute the query
```{r execute_dbplyr_query}
mtcars3 %>% collect()
```

plot the subset of mtcars between 3 and 4 (inclusive)
```{r plot-mtcars-between-3-and-4}
ggplot(mtcars3) +
  geom_point(aes(x=wt, y=mpg))
```

Make a query to the database using SQL. Since our dataset is small, we can use the [dbGetQuery()](https://dbi.r-dbi.org/reference/dbgetquery) function to send the query to the database, retrieve the results, then clean the result set, all in a single command:
```{r query-database-with-sql-small-dataset}
mtcars4 <- dbGetQuery(con, "SELECT * FROM `mtcars` WHERE ((`wt` >= 2) AND (`wt` <= 3))")
mtcars4
```


plot the subset of mtcars between 2 and 3 (inclusive)
```{r plot-mtcars-between-2-and-3}
ggplot(mtcars4) +
  geom_point(aes(x=wt, y=mpg))
```

If the dataset is larger, we should switch the the [dbSendQuery()](https://dbi.r-dbi.org/reference/dbsendquery), [dbFetch()](https://dbi.r-dbi.org/reference/dbfetch), and [dbClearResult()](https://dbi.r-dbi.org/reference/dbclearresult) functions to execute the query, retrieve the results, and clear the result set. The `dbFetch()` function takes a parameter, `n`, that allows you to control the number of results that are returned.

```{r query-database-with-sql-large-dataset}
res <- dbSendQuery(con, "SELECT * FROM `mtcars` WHERE ((`wt` >= 2) AND (`wt` <= 3))")
mtcars5 <- dbFetch(res, n=8)
dbClearResult(res)
mtcars5
```


disconnect the database connection:
```{r}
dbDisconnect(con)
```

Helpful websites:
1. [RSQLite Documentation](https://rsqlite.r-dbi.org/)
2. [Writing a dataframe to a database](https://www.datacareer.de/blog/connect-to-postgresql-with-r-a-step-by-step-example/)
3. [dbplyr Documentation](https://dbplyr.tidyverse.org)
4. [Make a ggplot2 plot](http://michaeltoth.me/a-detailed-guide-to-the-ggplot-scatter-plot-in-r.html)
5. [DBI Tutorial](https://dbi.r-dbi.org/articles/dbi)